Supplementary Information

### 2021-06-09 doc/manuscript/ForestGEO_treerings.Rmd
Google docs changes (results, including figures)

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each. The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the most influential window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

@@

```
At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential seasonal windows. 
However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics.
For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig. `r climwin_SCBI_W`). 
For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., PET at SCBI Fig. `r climwin_SCBI_T`), 
and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Fig. `r climwin_HF_W`).
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).

@@

Within the temperature group (Fig. `r schematic`), the most commonly selected variables were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites. 
$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
The most influential seasonal windows for temperature tended to be shorter than those of precipitation ($\le$ 3 months at 9 of 10 sites).
They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).

@@

![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 5), relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

```{r , eval=TRUE, echo=FALSE}
###other:
n_ssc= 45 #number species-site combinations
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
percent_T_or_P_responses_nonlinear=(32+18)/(42+38)
```
Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species-site combinations, and were predominantly positive (Table `r hypotheses`, Fig. `r clim_sens`).
Specifically, there were positive first-order linear terms for precipitation all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig. `r clim_sens`).

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with the most influential seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution. Other terms in the model are held constant at their median. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)

@@

When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs. `r clim_sens`, `r GLS_BCNM`).
In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r clim_sens`, `r size`).
The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig. `r GLS_SCBI`), but exceptions occurred when RW decreased steeply with DBH (e.g., at LT, Fig. `r GLS_LT`).

@@

Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, typically increasing to a peak at intermediate DBH and then declining (Table `r hypotheses`, `Fig. `r size`).
Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
For $\Delta$AGB, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).

![**Figure 5 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model. Other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in top model. Other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. ](tables_figures/Year_responses_BAI_only.png)

>>>>

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each. The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the most influential window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

```
At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response, and relative strength of correlation, to climate variables across the range of potential seasonal windows. 
However, the seasonal window exhibiting the strongest climatic effect on growth, and even the most influential climate variable, sometimes differed among the growth metrics.
For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig. `r climwin_SCBI_W`). 
For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., PET at SCBI Fig. `r climwin_SCBI_T`), 
and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Fig. `r climwin_HF_W`).
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).

@@

Within the temperature group (Fig. `r schematic`), the most commonly selected variables were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites. 
$T_{min}$ was identified as the top driver at BCNM, where its effects were only marginally significant for one species (Fig. `r clim_sens`). 
$T_{mean}$ was never selected as the top driver.
The most influential seasonal windows for temperature tended to be shorter than those of precipitation ($\le$ 3 months at 9 of 10 sites).
They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).

@@

![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 5), relationships are plotted if included in the top model. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

```{r , eval=TRUE, echo=FALSE}
###other:
n_ssc= 45 #number species-site combinations
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
percent_T_or_P_responses_nonlinear=(32+18)/(42+38)
```
Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species-site combinations, and were predominantly positive (Table `r hypotheses`, Fig. `r clim_sens`).
Specifically, there were positive first-order linear terms for precipitation for all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig. `r clim_sens`).

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with the most influential seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominantly positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)

@@

When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were selected for the top model, regardless of the growth metric used.
Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs. `r clim_sens`, `r GLS_BCNM`).
In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r clim_sens`, `r size`).
The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig. `r GLS_SCBI`), but exceptions occurred when RW decreased steeply with DBH (e.g., at LT, Fig. `r GLS_LT`).

@@

Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, typically increasing to a peak at intermediate DBH and then declining (Table `r hypotheses`, Fig. `r size`).
Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
For $\Delta$AGB, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).

![**Figure 5 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. ](tables_figures/Year_responses_BAI_only.png)

### 2021-06-09 doc/manuscript/ForestGEO_treerings_SI.Rmd
Google docs changes (results, including figures)

GLS_generic_legend = " for all three growth metrics examined here. For each species that met the criteria for inclusion in the analysis, relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD. Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/composite_plots."

@@

![`r SC_1970break_legend` *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. For each relationship shown, other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }

@@

![`r RW_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the DBH range. Other terms in the model are held constant at their median. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }

>>>>

GLS_generic_legend = " for all three growth metrics examined here. For each species that met the criteria for inclusion in the analysis, relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/composite_plots."

@@

![`r SC_1970break_legend` *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }

@@

![`r RW_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the DBH range. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
